Storm surges in the Strait of Georgia simulated with a regional model

### 2015-06-08 AO_guides/manuscriptAO.tex
more typos

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. {\color{red} At the north, seasonal climatologies from \citet{thomson1981oceanography} were used.} %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. A harmonic analysis was performed for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series was first filtered by applying a Doodson filter \citep{parker2007NOAA} and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc were not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

@@

In general, the K$_1$ tidal harmonics are well-represented by the model with complex differences below 5 cm at all stations except Seymour Narrows (station 27). Seymour Narrows is a very constricted region in Johnstone Strait and is not well resolved by the model. As such, the M$_2$ amplitude and phase are also poorly reproduced here.  The model's M$_2$ amphidrome, indicated by the drop in amplitude around station 5 (Victoria), is too strong resulting in lower than observed M$_2$ amplitudes in this region. {\color{red} The model also represents the dip in M$_2$ amplitude around stations 25 and 26 reasonably well.} Additionally, the M$_2$ phase errors in the Strait of Juan de Fuca (stations 1-7) are quite large, leading to large complex differences in this region.  The match in the Strait of Georgia (stations 10-24) is very good, with complex differences less than 6 cm, in part due to tidal tuning for highly accurate tides in the main region of interest.  {\color{red}Because the tides were tuned for a good match in the Strait of Georgia, the match in the Strait of Juan de Fuca has suffered, particularly noticeable at the station closest to the open boundary (Port Renfrew, 1). } Details on the tidal tuning procedure are provided in Appendix \ref{sec:appendix}.

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error. Note that the model will contain tidal energy from shallow water constituents due nonlinear interactions in the model equations. As such, the correction is calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (SA, SSA, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model sea surface height as a correction. Note that the harmonic analysis used to generate the constituents is applied in the same manner as that used in the forcing. That is, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy and then the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals, the $\gamma^2$ measure should be much lower for the total water levels; hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.  {\color{red} Observed and modelled maximum water levels and surge values as well as differences in their timing are reported in Table \ref{tab:peak}.}

@@

{\color{red} This case was also used to asses the importance of the northern boundary residual forcing and was used to brielfy examine stratified effects. When the northern boundary forcing is exlcuded the surge amplitude at Campbell River drops by about 10 cm. The other stations were not very much affected. This was also observed in another hindcast (Dec 2006). The effect of stratification was examined by considering a case with an initially horizontally uniform stratification and comparing to the fully stratified case. The largest differences were observed at Victoria, with an average 5 cm increase in sea surface height after an initial adjustment period. Other studies have seen an improvement in modelled surge amplitudes when using a baroclinic model as opposed to a three-dimensional barotropic model \citep{ma2015oceanic}. We plan to investigate stratified effects in more detail in a future study.}

@@

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, local atmospheric forcing (both local winds and local atmospheric pressure), and sea surface height anomaly at the open boundary (also called remote forcing) in the black curve. Second, a simulation with all forcing except the local atmospheric forcing in the dashed green curve. Third, a simulation with all forcing except the remote forcing in the red curve. Fourth, a simulation with only tides and local wind forcing, that is, excluding remote forcing and local atmospheric pressure forcing in the dashed blue curve. {\color{red} Comparisons between these simulations will determine the contributions from the remote forcing, the local wind forcing, and the local atmospheric pressure forcing.} Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction (Figure \ref{fig:tidesurge}).

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above {\color{red} 0.84} at these four locations. {\color{red} There is considerable noise in the observed Campbell River residual, which could explain its low skill score (Table \ref{tab:statistics}.} {\color{red} The modelled residuals in each simulation deviate only slightly from each other, most noticeable at Point Atkinson during the strong wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher.  The maximum residuals on Dec 17, 2012 are almost identical at each station (Table \ref{tab:peak}).} Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was {\color{red}66.3 cm}. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmospheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain; however, local wind patterns contribute to wave conditions which can impact flooding and are not accounted for in this model.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge are not accounted for in the model. Since the remote forcing is an important contributing factor, any forecasting with this regional model must make use of results from a larger scale prediction system.

>>>>

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. {\color{red} At the north, seasonal climatologies from \citet{thomson1981oceanography} were used in the same manner.} %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with the tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. A harmonic analysis was performed for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series was first filtered by applying a Doodson filter \citep{parker2007NOAA} and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc were not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

@@

In general, the K$_1$ tidal harmonics are well-represented by the model with complex differences below 5 cm at all stations except Seymour Narrows (station 27). Seymour Narrows is a very constricted region in Johnstone Strait and is not well resolved by the model. As such, the M$_2$ amplitude and phase are also poorly reproduced here.  The model's M$_2$ amphidrome, indicated by the drop in amplitude around station 5 (Victoria), is too strong resulting in lower than observed M$_2$ amplitudes in this region. {\color{red} The model also represents the dip in M$_2$ amplitude around stations 25 and 26 reasonably well.} Additionally, the M$_2$ phase errors in the Strait of Juan de Fuca (stations 1-7) are quite large, leading to large complex differences in this region.  The match in the Strait of Georgia (stations 10-24) is very good, with complex differences less than 6 cm, in part due to tidal tuning for highly accurate tides in the main region of interest.  {\color{red}Because the tides were tuned for a good match in the Strait of Georgia, the match in the Strait of Juan de Fuca was compromised, particularly noticeable at the station closest to the open boundary (Port Renfrew, 1). } Details on the tidal tuning procedure are provided in Appendix \ref{sec:appendix}.

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error. Note that the model will contain tidal energy from shallow water constituents due to nonlinear interactions in the model equations. As such, the correction is calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (SA, SSA, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model sea surface height as a correction. Note that the harmonic analysis used to generate the constituents is applied in the same manner as that used in the forcing. That is, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy and then the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one and values close to one indicate good model skill. Since the Willmott skill score is nondimensional, it is an appropriate statistic to report when comparing skills of different variables, in this case, the total water level and the residuals. Since the total water level has a larger variability than the residuals, the $\gamma^2$ measure should be much lower for the total water levels; hence, the Willmott skill score provides an additional point of comparison. These three statistics, along with the mean observed values and mean modelled values (which indicate the model's bias) are reported in Table \ref{tab:statistics} for each hindcast.  {\color{red} Observed and modelled maximum water levels and surge values as well as differences in their timing are reported in Table \ref{tab:peak}.}

@@

{\color{red} This case was also used to assess the importance of the northern boundary residual forcing and was used to briefly examine stratified effects. When the northern boundary forcing is excluded the surge amplitude at Campbell River drops by about 10 cm. The other stations were not strongly affected. This was also observed in another hindcast (Dec 2006). The effect of stratification was examined by considering a case with an initially horizontally uniform stratification and comparing to the fully stratified case. The largest differences were observed at Victoria, with an average 5 cm increase in sea surface height after an initial adjustment period. Other studies have seen an improvement in modelled surge amplitudes when using a baroclinic model as opposed to a three-dimensional barotropic model \citep{ma2015oceanic}. We plan to investigate stratified effects in more detail in a future study.}

@@

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, local atmospheric forcing (both local winds and local atmospheric pressure), and sea surface height anomaly at the open boundary (also called remote forcing) in the black curve. Second, a simulation with all forcing except the local atmospheric forcing in the dashed green curve. Third, a simulation with all forcing except the remote forcing in the red curve. Fourth, a simulation with only tides and local wind forcing, that is, excluding remote forcing and local atmospheric pressure forcing in the dashed blue curve. {\color{red} Comparisons between these simulations will determine the contributions from the remote forcing, the local wind forcing, and the local atmospheric pressure forcing.} Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction (Figure \ref{fig:tidesurge}).  {\color{red} Although the winds and atmospheric pressure are influenced by large scale atmospheric features, we use the terms local and remote to differentiate between factors originating from inside and outside of the domain.}

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above {\color{red} 0.84} at these four locations. {\color{red} There is considerable noise in the observed Campbell River residual, which could explain its low skill score (Table \ref{tab:statistics}).} {\color{red} The modelled residuals in each simulation deviate only slightly from each other, most noticeable at Point Atkinson during the strong wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher.  The maximum residuals on Dec 17, 2012 are almost identical at each station (Table \ref{tab:peak}).} Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was {\color{red}66.3 cm}. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmospheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain; however, local wind patterns contribute to wave conditions which can impact flooding and are not accounted for in this model.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge is not accounted for in the model. Since the remote forcing is an important contributing factor, any forecasting with this regional model must make use of results from a larger scale prediction system.
